---
title: "Companion SARA Prioritization Model Document"
author: "Chris Madsen & John Phelan"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)

library(readxl)
library(tidyverse)
library(leaflet)
library(leafpop)
library(sf)
library(RColorBrewer)
library(patchwork)

```

```{r run_data_prep_scripts}
source("scripts/utils/load_data.R")
source("scripts/utils/prep_predictor_data_f.R")
```

## SARA Prioritization Model

### Objective
Identify the overlap between federally listed Species-at-Risk species in the Columbia and Fraser watersheds and aquatic invasive species. Estimate the potential impact of AIS on SAR.

### Study Area
These two spatial layers are trimmed to just the area inside the Fraserand Columbia River Basins (the purple area in the figure below).

```{r show_fraser_and_columbia_basins}
ggplot() + 
  geom_sf(data = sf::st_as_sf(bc), fill = 'lightgrey') + 
  geom_sf(data = frascol, fill = 'purple') + 
  ggthemes::theme_map() + 
  ggspatial::annotation_scale()
```

### Datasets

This model is based on spatial overlap of polygons maintained by the
Department of Fisheries and Oceans (DFO) that represent known
occurrences of SARA-listed species (or SAR, i.e., species-at-risk), and occurrence records of aquatic invasive species (AIS) in British
Columbia. The overlap is found based on named rivers, streams, and lakes within the BC Freshwater Atlas. The locations of the SAR are represented by polygons (and points for hte Sockeye salmon in Cultus Lake).

#### SARA Locations {.tabset}
```{r plot_sar_locations}

species_list <- c("Cultus Pygmy Sculpin", "Westslope Cutthroat Trout", 
                  "Bull Trout", "Rocky Mountain Ridged Mussel", 
                  "Sockeye Salmon")
my_palette <- brewer.pal(n = 5, name = "Dark2")
species_colours <- setNames(my_palette, species_list)

# Try making some plots!
dfo_sar_4326 = sf::st_transform(dfo_sar, 4326)

dfo_extent = sf::st_bbox(dfo_sar_4326) |>
  sf::st_as_sfc() |>
  sf::st_sf() |>
  sf::st_set_crs(4326)

species_maps = dfo_sar_4326 |>
  dplyr::group_by(Common_Name_EN) |>
  dplyr::group_split() |>
  purrr::map( ~ {

    buffered_bbox = sf::st_bbox(.x) |>
      sf::st_as_sfc() |>
      sf::st_as_sf() |>
      sf::st_transform(3005) |>
      sf::st_buffer(dist = 500)

    the_ocean_bm = basemaps::basemap_terra(ext = buffered_bbox,
                                           map_service = "osm",
                                           map_type = "streets")

    ggplot() +
      tidyterra::geom_spatraster_rgb(data = the_ocean_bm) +
      geom_sf(data = sf::st_simplify(.x),
              aes(fill = Common_Name_EN), alpha = 0.8, linewidth = 0.8) +
      scale_fill_manual(values = species_colours) +  # 👈 Custom fill colours
      ggtitle(case_when(
        .x$Common_Name_EN[1] == "Cultus Pygmy Sculpin" ~ "Coastrange Sculpin - Cultus",
        .x$Common_Name_EN[1] == "Westslope Cutthroat Trout" ~ "Westslope Cutthroat Trout",
        .x$Common_Name_EN[1] == "Bull Trout" ~ "Bull Trout - South Coast",
        .x$Common_Name_EN[1] == "Rocky Mountain Ridged Mussel" ~ "Rocky Mountain Ridged Mussel",
        .x$Common_Name_EN[1] == "Sockeye Salmon" ~ "Sockeye Salmon - Cultus",
        TRUE ~ .x$Common_Name_EN[1]
      )) +
      ggthemes::theme_map() +
      theme(legend.position = 'none',
            plot.title = element_text(hjust = 0.5, size = 12, face = "bold"))
  })

all_my_maps = Reduce(`+`, species_maps)


inset_map = ggplot() +
  geom_sf(data = bcmaps::bc_bound(), fill = "white") +
  geom_sf(data = dfo_extent, color = "red", fill = NA, linewidth = 1) +
  ggthemes::theme_map()

all_my_maps + inset_map

```



#### Named waterbodies 
Within the Columbia and Fraser watersheds, there are 4477 named waterbodies
```{r plot_named_waterbodies}

#combine the fras and colum wat
named_wbs_fras_col = named_wbs |>
  sf::st_intersection(frascol)

# Plot the named waterbodies
ggplot()+
  geom_sf(data = st_as_sf(bc))+
  geom_sf(data = named_wbs_fras_col, fill = "steelblue", alpha = 0.5) +
  geom_sf(data = frascol, fill = NA, color = "black", linewidth = 0.75) +
  ggthemes::theme_map()

```

#### Filtered waterbodies by SARA locations
Filtering the named waterbodies to only those that intersect with the SARA locations. This results in 720 named waterbodies with the listed SAR species in the Columbia and Fraser watersheds.
```{r filter_named_waterbodies_plot}

sf::sf_use_s2(FALSE)

filtered_wbs <- named_wbs_fras_col[sf::st_intersects(named_wbs_fras_col, dfo_sar, sparse = FALSE) |> rowSums() > 0, ]

ggplot()+
  geom_sf(data = st_as_sf(bc))+
  geom_sf(data = filtered_wbs, fill = "steelblue", alpha = 0.5) +
  geom_sf(data = frascol, fill = NA, color = "black", linewidth = 0.75) +
  ggthemes::theme_map()

sf::sf_use_s2(TRUE)
```




-   Risk assessments ?? Alert sheets?

## Model Process

### SAR-AIS overlap

Both SAR and AIS are spatially joined to named waterbodies (i.e. lakes,
rivers and streams) to identify areas of overlap. In addition, each
instance of overlap between SAR and AIS is also tested for the presence
of AIS in the waterbody immediately upstream. Note that if the upstream
waterbody was a river or stream and there was one or more lakes in
physical proximity, the lake(s) was/were also included in the scope of
this 'upstream waterbody' analysis.

-   Add occurrence dots for AIS? Maybe also polygon(s) for DFO SARA in
    WB?

```{r upstream_example}
eg_wb = named_wbs[named_wbs$waterbody == 'Alouette River',]
row_name = "Alouette River"
row_fwa = eg_wb$FWA_WATERSHED_CODE
row_fwa_prefix = str_remove(row_fwa, "000000.*")
row_watershed = eg_wb$watershed
eg_wb = eg_wb |> 
  dplyr::mutate(id = "Alouette River")
# Filter named waterbodies for just those upstream of the_wb.
# Then, do AIS overlap with just those upstream wb(s).
upstream_named_wbs = named_wbs |>
  dplyr::filter(str_detect(FWA_WATERSHED_CODE,paste0(row_fwa_prefix))) |>
  dplyr::filter(FWA_WATERSHED_CODE != row_fwa) |>
  dplyr::filter(str_detect(FWA_WATERSHED_CODE,paste0(row_fwa_prefix,"(?!000000)"))) |>
  dplyr::filter(str_detect(FWA_WATERSHED_CODE,paste0(row_fwa_prefix,"[0-9]{6}-000000"))) |> 
  dplyr::mutate(id = "Lakes/Rivers \none level \nupstream")
  # sf::st_drop_geometry()

# Snag connecting streams. We don't actually do this in the model, but this might be necessary for illustrative purposes.
fwa_pattern = bcdata::CQL(paste0("FWA_WATERSHED_CODE like '",row_fwa_prefix,"%'"))

rel_streams = bcdata::bcdc_query_geodata("freshwater-atlas-stream-network") |> 
  filter(fwa_pattern) |> 
  collect() |> 
  dplyr::summarise() |> 
  dplyr::mutate(id = "Unnamed connecting streams")

ggplot() + 
  geom_sf(data = rel_streams, aes(col = id)) +
  geom_sf(data = eg_wb, aes(fill = id, col = id)) + 
  geom_sf(data = upstream_named_wbs, aes(fill = id, col = id)) + 
  scale_color_manual(values = c("Alouette River" = "purple",
                                "Lakes/Rivers \none level \nupstream" = "darkblue",
                                "Unnamed connecting streams" = "lightblue")) + 
  guides(fill = "none") + 
  labs(color = "Waterbody") + 
  ggspatial::annotation_scale(pad_x = unit(7.5,'cm')) +
  ggthemes::theme_map() +
  theme(legend.position = 'left')
```

### Strength of Potential Negative Effects of AIS on SAR

Species-specific interactions between AIS and SAR were collated from
available risk assessments and AIS 'Alert Sheets' (informative bulletins
prepared by professional biologists to inform the public of aquatic
invasive species in British Columbia). Each AIS was rated on its ability
to negatively affect SARA fish and SARA invertebrates through:

-   predation

-   competition

-   habitat modification

-   nutrient depletion

-   disease/parasite transmission

The uncertainty of the strength of these effects was also assessed.

```         
**Maybe mention that we took the average of those columns.**
```

For each waterbody with SAR / AIS overlap, or SAR present with AIS
present in the nearby upstream, the most severe potential effect of AIS
on SAR was identified to represent the overall risk to SAR in that
waterbody.

### COSEWIC status

The COSEWIC status of each SAR also contributed to the overall risk at
the waterbody level: a status of "Not at Risk" did not raise the overall
risk estimate, a status of either "Data Deficient" or "Special Concern"
added 1 to the risk estimate, a status of "Threatened" added 2 to the
risk estimate, and a status of "Endangered" added 3 to the risk
estimate.

### Reduction for AIS upstream

The maximum potential negative effect score and COSEWIC status score of
AIS that were upstream of a given waterbody with SAR present was reduced
by half. This was intended to attenuate the estimated risk when AIS were
less proximal.

## Final Equation

$$
Risk_{WB} = (E_{WB} + S_{WB}) + \frac{(E_{U} + S_{U})}{2} 
$$

Where $Risk_{WB}$ is the overall estimate risk rating of a waterbody,
$E$ is the maximum potential negative effect of the AIS present in the
waterbody on the SAR, and $S$ is the COSEWIC status of the SAR. The
index $_{WB}$ refers to the waterbody in question, and the index $_{U}$
refers to the named rivers or lakes that are one 'step' upstream from
the waterbody in question, according to the Freshwater Atlas.

```{r map_of_the_province_with_ratings_raster}
# Read in results excel file.
result_files = list.files(path = 'output/', 
                          pattern = '^SARA.*.xlsx$',
                          full.names = T)
# Figure out which file is the most recent.
file_to_read = result_files |> 
  lapply(\(x){
    data.frame(filename = x, c_time = file.info(x)$ctime)
  }) |> 
  dplyr::bind_rows() |> 
  dplyr::arrange(dplyr::desc(c_time)) |> 
  dplyr::slice(1)

results = openxlsx::read.xlsx(file_to_read$filename)

results = results |> 
  dplyr::select(waterbody, watershed, Common_Name_EN, Population_EN,
                final_risk, final_risk_b)

# Make sure we don't have NA values for lowest final risk bin.
results = results |> 
  dplyr::mutate(final_risk_b = tidyr::replace_na(final_risk_b, "1"))

rel_named_wbs = named_wbs |> 
  dplyr::filter(waterbody %in% results$waterbody)
  
rel_named_wbs = rel_named_wbs |> 
  dplyr::group_by(waterbody, watershed) |> 
  dplyr::summarise() |> 
  dplyr::ungroup()

rel_named_wbs_w_results = rel_named_wbs |> 
  dplyr::inner_join(results)

bbox_of_results = sf::st_bbox(frascol)

rel_ecos = bcmaps::ecosections() |> 
  sf::st_transform(4326) |> 
  sf::st_filter(frascol)
  
# Try to find average or sum of final risk bins, display either as polygons or raster?
# for each ecosection, find the sum of the final risk bin where there is 
# spatial overlap
joined_ecos_scores <- sf::st_join(rel_ecos, rel_named_wbs_w_results, join = sf::st_intersects, left = TRUE)

risk_polys <- joined_ecos_scores |>
  mutate(final_risk_b = as.numeric(final_risk_b)) |>
  group_by(ECOSECTION_NAME) |>
  summarise(
    sum_final_risk_b = sum(final_risk_b, na.rm = TRUE),
    Common_Names = paste(unique(na.omit(Common_Name_EN)), collapse = ", "),
    Populations = paste(unique(na.omit(Population_EN)), collapse = ", ")
  ) |>
  mutate(sum_final_risk_b = ifelse(is.na(sum_final_risk_b), 0, sum_final_risk_b)) |>
  ungroup()
  


# rel_ecos = rmapshaper::ms_simplify()
# ggplot() +
#   geom_sf(data = bcmaps::ecosections())

ggplot() + 
  geom_sf(data = sf::st_as_sf(bc), col = 'lightgrey', fill = 'transparent') + 
  geom_sf(data = frascol, col = 'black') +
  geom_sf(data = rel_named_wbs_w_results, aes(fill = final_risk_b, col = final_risk_b)) + 
  coord_sf(xlim = bbox_of_results[c(1,3)],
           ylim = bbox_of_results[c(2,4)]) + 
  ggthemes::theme_map() + 
  theme(legend.position = 'left') + 
  labs(fill = "Risk Estimate", col = "Risk Estimate") + 
  ggspatial::annotation_scale()

# Maybe make the above map a leaflet map? Or add that?




leaflet::leaflet(risk_polys) |> 
  leaflet::addTiles() |> 
  leaflet::addPolygons(
    fillColor = ~leaflet::colorNumeric("viridis", sum_final_risk_b)(sum_final_risk_b),
    color = "black",
    weight = 1,
    opacity = 1,
    fillOpacity = 0.7,
    popup = leafpop::popupTable(risk_polys, 
                                zcol = c("Common_Names", "Populations", "sum_final_risk_b")),
    highlightOptions = leaflet::highlightOptions(
      weight = 2,
      color = "#666",
      fillOpacity = 0.7,
      bringToFront = TRUE
    )
  ) |> 
  leaflet::addLegend(
    pal = leaflet::colorNumeric("viridis", risk_polys$sum_final_risk_b),
    values = risk_polys$sum_final_risk_b,
    position = "bottomright"
  )

```

## Future Work

### MaxEnt Assessment of Habitat Suitability

Modelling habitat suitability for AIS across the watersheds, gives an
indication as to how suitable a particular waterbody would be for an
AIS. It takes into account a number of environmental variables, such as
temperature and calcium concentration. The output may be combined with
current risk score, for AIS upstream of a SARA listed species. The
usability is uncertain as to how the output of MaxEnt would be
incorporated in the current modelling process, with there being a
"suitable/unsuitable" binary distinction in habitat suitability.

### First Nations Cultural Significance and Priority

The cultural significance of a SARA listed species should be
incorporated to ensure that these species in waterbodies are noted to be
offered further protection, or maybe suitable for more detailed analysis
and action. The inclusion of knowledge from first nations would make a
valuable contribution to the assessment of the SARA listed species in
the waterbodies.
